---
format: html
---

# Banking Fraud Behaviour and Risk Analysis Report

### Project Objective

This project examines banking transaction behaviour to understand where fraud risk may concentrate across customer activity, payment channels, authentication methods, account maturity, and transaction monitoring signals. The analysis is designed to help banking stakeholders move from a general view of transactions to a more targeted understanding of which behavioural patterns may require closer operational and fraud-risk attention.

The business problem being investigated is whether transaction-level attributes can reveal meaningful differences between lower-risk and higher-risk activity. In banking, fraud rarely depends on a single field in isolation. Transaction amount, channel usage, device risk, anomaly behaviour, authentication type, international activity, failed activity, account age, and geographic distance can all contribute to a broader risk profile.

Understanding these patterns matters because financial institutions must balance fraud prevention with customer experience. Stronger monitoring can reduce loss exposure, but unnecessary friction can harm legitimate customers. A structured transaction analysis helps identify where controls, monitoring, and resources should be focused while preserving efficient service for routine banking activity.

### Key Findings

- Transaction value behaviour provides an important baseline for understanding normal activity, high-value transaction exposure, and potential concentration in the upper end of the transaction distribution.

- Payment channel usage highlights how customers interact with banking services and can help identify where operational workload, digital service demand, and transaction value are concentrated.

- Risk indicators such as device risk score, anomaly score, transaction velocity, suspicious IP activity, and fraud flags create a stronger view of fraud exposure when assessed alongside transaction behaviour rather than in isolation.

- Time-based transaction patterns can reveal periods of heavier operational demand and potential monitoring pressure, supporting better alignment of fraud review capacity and transaction oversight.

- Authentication behaviour and account security signals, including login attempts and failed transactions, provide useful evidence for identifying behavioural stress patterns that may be associated with fraud risk.

- Account age and combined behavioural interactions suggest that fraud risk should be assessed across the customer lifecycle and across combinations of behaviours, not only through single-variable rules.

### Business Impact

The findings can support fraud prevention by helping risk teams identify transaction profiles that deserve closer monitoring, such as combinations of high risk scores, suspicious behaviour, international activity, authentication patterns, and unusual transaction velocity. This creates a foundation for more targeted fraud detection rules and more efficient investigation queues.

For operations teams, the analysis supports workload planning by showing where transaction activity may concentrate across channels, time periods, and higher-risk segments. This can help align monitoring coverage, staffing, and escalation processes with the areas most likely to require attention.

For customer risk management and authentication strategy, the analysis provides evidence for differentiating controls by behaviour and lifecycle stage. Newer accounts, higher-risk authentication journeys, and transactions with multiple risk signals may require stronger controls, while lower-risk activity can remain smoother for customers.

Overall, this work provides a practical analytical foundation for future fraud detection initiatives, including risk scoring, alert prioritisation, segmentation, and targeted monitoring strategies. It helps banking leaders connect transaction behaviour to fraud risk in a way that is commercially useful, operationally actionable, and aligned with customer experience considerations.

## Dataset Overview

### About the Dataset

The dataset represents individual banking transactions and the surrounding behavioural signals that may help explain transaction risk. Each record reflects a transaction event, including the value of the transaction, how the transaction was initiated, the security context around the activity, and whether fraud-related indicators were present.

The information available supports a broad view of banking behaviour. It includes transaction activity such as transaction value, payment channel, transaction timing, and transaction frequency. It also includes customer behaviour and account context, such as account maturity, login activity, recent failed transactions, and transaction velocity. From a fraud and security perspective, the dataset includes authentication method, card presence, international transaction activity, device risk, anomaly behaviour, suspicious IP activity, and fraud status.

This combination of transaction, behavioural, authentication, and risk information makes the dataset suitable for analysing how fraud risk emerges across different customer journeys. Rather than treating fraud as a single outcome, the dataset allows the analysis to explore patterns across transaction value, channel usage, account lifecycle stage, authentication behaviour, geographic context, and interacting risk signals.

### Business Context

Transaction monitoring is a core capability in modern banking because fraud risk can emerge quickly, across multiple channels, and through increasingly complex customer behaviours. Banks must identify suspicious activity early enough to prevent financial loss, protect customers, and meet risk management expectations, while still allowing legitimate transactions to move through the system with minimal friction.

One of the key challenges for banks is balancing control with customer experience. Overly strict monitoring can increase false positives, delay legitimate transactions, and create frustration for customers. Weak monitoring, however, can expose the institution and its customers to avoidable fraud losses. Effective transaction analytics helps banks identify where stronger intervention is justified and where routine activity can remain efficient.

By analysing transaction behaviour, authentication patterns, risk indicators, and lifecycle signals together, banking teams can make more informed operational decisions. These insights can support fraud prevention, monitoring prioritisation, customer protection, authentication policy design, and resource allocation across fraud operations, digital banking, and risk management teams.

### Dataset Scope

| Area | Summary |
|---|---|
| Industry | Retail banking and digital transaction monitoring |
| Analytical focus | Fraud risk, transaction behaviour, authentication patterns, and operational monitoring signals |
| Unit of analysis | Individual banking transaction |
| Observational level | Transaction-level activity with associated behavioural, security, and risk attributes |
| Type of data available | Transaction values, payment channels, account maturity, authentication behaviour, risk scores, fraud indicators, geographic context, and activity timing |

### Key Information Available

| Category | Information available | Contribution to fraud analysis and decision-making |
|---|---|---|
| Transaction Behaviour | Transaction values, transaction frequency, transaction velocity, and activity timing are available to describe how banking activity occurs across the transaction base. | Helps identify unusual transaction patterns, high-value exposure, operational workload peaks, and behaviours that may require closer monitoring. |
| Payment Activity | Payment channel and card presence information show how customers interact with banking services across digital, physical, and card-based transaction contexts. | Supports channel-level risk assessment, operational planning, and understanding whether certain transaction routes carry different fraud exposure. |
| Risk Indicators | Device risk, anomaly behaviour, fraud status, and suspicious IP signals provide direct evidence of transaction risk conditions. | Enables fraud teams to compare risk patterns across transaction types, identify elevated-risk segments, and prioritise monitoring attention. |
| Authentication & Security | Authentication method, login attempts, and recent failed transaction activity provide insight into security behaviour surrounding transactions. | Helps assess whether certain authentication journeys or behavioural stress signals are associated with higher fraud risk. |
| Customer Lifecycle | Account age provides a view of account maturity and supports comparison between newer and more established accounts. | Supports onboarding risk assessment and helps determine whether early account stages require stronger monitoring or controls. |
| Geographic & Contextual Factors | Geographic distance and international transaction indicators provide context on where and how transactions occur. | Helps identify whether international or geographically unusual activity shows different fraud, risk, or monitoring requirements. |

## Business Case-study

### Introduction

This analysis is structured around key business questions designed to understand transaction behaviour, operational activity patterns, fraud risk signals, authentication behaviour, customer lifecycle risk, and interaction effects between behavioural and fraud indicators. Together, these questions frame the project as a banking fraud and transaction monitoring use case focused on identifying where risk may emerge and how monitoring strategies can be better aligned to customer activity.

## Business Questions

### Theme 1: Customer Transaction Behaviour

These questions establish the baseline for understanding how customers transact and which behaviours represent normal transaction activity across the banking portfolio.

1. How are transaction amounts distributed across the bank's transaction portfolio, and what does normal customer spending behaviour look like?

2. Which payment channels are used most frequently, and how is transaction activity distributed across channels?

### Theme 2: Operational Activity Patterns

These questions focus on how transaction activity changes across risk levels and time periods, helping identify when operational teams may need stronger monitoring coverage or review capacity.

3. How do transaction values and customer behaviours vary across different levels of transaction risk?

4. When does transaction activity occur, and are there periods of elevated operational or fraud risk?

### Theme 3: Fraud Exposure & Contextual Risk

These questions examine contextual drivers of fraud exposure, including transaction geography, authentication behaviour, and account maturity.

5. Do international transactions exhibit different behavioural and fraud patterns compared with domestic transactions?

6. How does authentication behaviour influence fraud exposure and security risk?

7. How does account maturity influence customer behaviour and fraud risk?

### Theme 4: Fraud Interaction Patterns

This question brings together multiple behavioural and risk indicators to understand whether fraud risk is strongest when signals occur in combination.

8. Which combinations of customer behaviours and risk indicators are most strongly associated with fraudulent transactions?

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(janitor)
library(scales)

banking_transactions <- read_csv(
  "data/banking_transactions.csv",
  show_col_types = FALSE
) %>%
  clean_names()
```

## Analysis 1: Transaction Value Behaviour

### Business Question

How are transaction amounts distributed across the bank's transaction portfolio, and what does normal customer spending behaviour look like?

### Why This Matters

Understanding transaction value distribution is essential for distinguishing normal customer behaviour from unusual activity that may require closer review. In banking, most transactions are expected to fall within a routine behavioural range, while unusually large or highly concentrated values may indicate elevated exposure or potential fraud risk.

This analysis helps establish a practical view of customer spending behaviour, supports the identification of potential transaction outliers, and provides a foundation for setting risk-based monitoring thresholds that are sensitive to abnormal value patterns without over-escalating ordinary customer activity.

```{r}
#| label: analysis-1-transaction-value-distribution
#| fig-width: 9
#| fig-height: 5

transaction_value_summary <- banking_transactions %>%
  summarise(
    median_transaction_value = median(transaction_amount, na.rm = TRUE)
  )

banking_transactions %>%
  ggplot(aes(x = transaction_amount)) +
  geom_density(
    fill = "#14B8A6",
    color = "#0F766E",
    linewidth = 1,
    alpha = 0.35
  ) +
  geom_vline(
    xintercept = transaction_value_summary$median_transaction_value,
    color = "#7C2D12",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  scale_x_continuous(labels = label_dollar()) +
  labs(
    title = "Transaction Value Concentration",
    subtitle = "Density curve shows where transaction values cluster across the portfolio",
    x = "Transaction value",
    y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#4B5563"),
    panel.grid.minor = element_blank()
  )
```

### Key Findings

- Transaction amounts are concentrated within a visible central range, indicating a typical band of customer transaction behaviour.

- The distribution shows a tail of higher-value transactions, suggesting that a smaller group of transactions carries greater financial exposure.

- The median marker provides a clear reference point for separating routine transaction values from higher-value activity.

- High-value transactions appear less common than routine transactions, making them important candidates for risk-sensitive monitoring logic.

### Insights

Normal transaction behaviour is represented by the main concentration of transaction values, where most customer activity occurs. Unusual behaviour begins where transactions move away from this central range into the higher-value tail, especially when those transactions are combined with other risk signals.

For fraud detection systems, this means transaction value should be used as a contextual risk signal rather than a standalone trigger. Higher-value activity may not be fraudulent on its own, but it can materially increase exposure when paired with suspicious behaviour, abnormal device signals, or unusual transaction context.

# Analysis 2: Payment Channel Usage

[To be filled in next step]

# Analysis 3: Transaction Risk Behaviour

[To be filled in next step]

# Analysis 4: Transaction Activity Over Time

[To be filled in next step]

# Analysis 5: International vs Domestic Risk

[To be filled in next step]

# Analysis 6: Authentication Behaviour

[To be filled in next step]

# Analysis 7: Account Age & Lifecycle Risk

[To be filled in next step]

# Analysis 8: Fraud Interaction Patterns

[To be filled in next step]

# Final Recommendations

[To be filled in next step]

# Conclusion

[To be filled in next step]
